Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-12-11 paper.tex
Rephrase
Typo
Reviewer B:
  Line 47-48: "a purely functional program~~s~~".
Typo
Reviewer B:
  Line 49: "non-performance-critical".

As intimidating as cryptographers are, the essence of modern symmetric
cryptographic primitives can in fact be distilled in a few paragraph. A
primitive can be understood as a purely functional programs. It takes
two inputs: the ``cleartext'' and a list of ``round subkeys''
(which are derived from a single key through a non-performance
critical process called ``key schedule'' that we shall ignore here).
The output is the ``ciphertext''. The cleartext, the ciphertext and
each individual round subkeys are processed as ``blocks'' of binary
data. The block size is of the order of a hundred bits (typically, 64, 128
or 256 bits).

@@

To the disciples of Reynolds and Girard, this signature has come to
mean ``parametric data container'': how would you go about asserting
that the type constructor \coqe!Rows.T! is not doing something
non-trivial with the type it is provided as an argument? In other
words, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
as a \coqe!Rows.T nat!? One could try and argue that Coq does not
allow pattern matching on types but it is rather unsavory to involve
the design of the whole programming language into this argument.

Instead, we require \coqe!Rows.T! to offer a \coqe!map! operation: it
constructively witnesses the fact that a \coqe!Rows.T A! is
(functionally) related to a \coqe!Rows.T B!, as long as we can
(functionally) relate \coqe!A! and \coqe!B!. Put otherwise, we ask
that \coqe!Rows.T!, and similarly \coqe!Cols.T! and \coqe!Slice8.T!,
come equipped with their free theorems~\citep{reynolds:parametricity,
  wadler:free-theorem}.

>>>>

As intimidating as cryptographers are, the essence of modern symmetric
cryptographic primitives can in fact be distilled in a few paragraph. A
primitive can be understood as a purely functional program. It takes
two inputs: the ``cleartext'' and a list of ``round subkeys''
(which are derived from a single key through a non-performance-critical process called ``key schedule'' that we shall ignore here).
The output is the ``ciphertext''. The cleartext, the ciphertext and
each individual round subkeys are processed as ``blocks'' of binary
data. The block size is of the order of a hundred bits (typically, 64, 128
or 256 bits).

@@

To the disciples of Reynolds and Girard, this signature has come to
mean ``parametric data container'': how would you go about asserting
that the type constructor \coqe!Rows.T! is not doing something
non-trivial with the type it is provided as an argument? In other
words, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
as a \coqe!Rows.T nat!? One could try and argue that Coq does not
allow pattern matching on types but it is rather unsavory to involve
the design of the whole programming language into this argument.
%
Instead, we require \coqe!Rows.T! to offer a \coqe!map! operation: this
constructively witnesses the fact that a \coqe!Rows.T A! is
(functionally) related to a \coqe!Rows.T B!, as long as we can
(functionally) relate \coqe!A! and \coqe!B!. Put otherwise, we ask
that \coqe!Rows.T!, and similarly \coqe!Cols.T! and \coqe!Slice8.T!,
come equipped with their free theorems~\citep{reynolds:parametricity,
  wadler:free-theorem}.
